---
title: 'RF -> eBird'
output:
  pdf_document: default
  html_notebook: default
  html_document: default
---

**You Will Need:**

* The current raster stack

* Your duck species training and test sets

* R packages: raster, randomForest

```{r}
source("Y:/users/rah/R/Rconfigure_default.r")
library(knitr)
library(raster)
library(randomForest)
knitr::opts_knit$set(root.dir = "Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599")
```

```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/trainData_Oxyura_jamaicensis.RData") # trainData
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/testData_Oxyura_jamaicensis.RData") # testData
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/testSubset_Oxyura_jamaicensis.RData") # testSubset
```

```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v4.RData") # envStack
```

# Prepare training data

```{r}
trainData = cbind(trainData,
                  extract(envStack,trainData[,c(4,3)]))
trainData$GlobCover2009LC = as.factor(trainData$GlobCover2009LC)
trainData$LC_5min_global_2012 = as.factor(trainData$LC_5min_global_2012)

# Remove datapoints with missing values
trainData = trainData[-which(rowSums(is.na(trainData[,c(8,12,13,14,15,17,23:45)]))>0),]

# Convert presAbs to factor
trainData$presAbs = as.factor(trainData$presAbs)
```

Subsample absence for computational considerations?
```{r}
trainData = rbind(trainData[which(trainData$presAbs==1),], # all presences
                  trainData[sample(which(trainData$presAbs==0),20000),]) # 20000 absences
```

# Tune parameters for presence-absence model
We can start with a baseline/default strategy here by using the OOB error rate to tune ntree and mtry. This run starts with the default mtry - we can also try doubling and halving the default. For each, we can keep increasing ntree until the error rate is not improving anymore. 
```{r}
set.seed(1234) 
model1 = randomForest(presAbs~Isothermality + 
                              Annual.Mean.Temperature + 
                              Mean.Diurnal.Range + 
                              Temperature.Seasonality + 
                              Max.Temperature.of.Warmest.Month + 
                              Min.Temperature.of.Coldest.Month + 
                              Temperature.Annual.Range + 
                              Mean.Temperature.of.Wettest.Quarter + 
                              Mean.Temperature.of.Driest.Quarter + 
                              Mean.Temperature.of.Warmest.Quarter + 
                              Mean.Temperature.of.Coldest.Quarter + 
                              Annual.Precipitation + 
                              Precipitation.of.Wettest.Month + 
                              Precipitation.of.Driest.Month + 
                              Precipitation.Seasonality + 
                              Precipitation.of.Wettest.Quarter + 
                              Precipitation.of.Driest.Quarter + 
                              Precipitation.of.Warmest.Quarter + 
                              Precipitation.of.Coldest.Quarter + 
                              elevation + 
                              glwd_3 + 
                              GlobCover2009LC + 
                              LC_5min_global_2012 + 
                              TIME + 
                              COUNT_TYPE + 
                              EFFORT_HRS + 
                              EFFORT_DISTANCE_KM + 
                              EFFORT_AREA_HA + 
                              NUMBER_OBSERVERS,
                      data=trainData,
                      mtry=5, # default with 29 predictors (floor(sqrt(29)))
                      ntree=2000,
                      importance=TRUE, # compute importance values
                      proximity=FALSE, # don't compute proximity values
                      keep.forest=FALSE) # speed things up by not keeping the trees - we can regrow later for the final model
print(model1)
plot(model1$err.rate[,1])
```

# Select a final presence-absence model

# Tune parameters for abundance model

# Select a final abundance model

