---
title: "eBird: Null models"
output: html_notebook
---

Set-up chunk.
```{r,message=FALSE}
source("Y:/users/rah/R/Rconfigure_default.r")
library(ggplot2)
library(knitr)
library(sp)
library(raster)
library(dismo)
library(PRROC)
library(pROC)
#library(tidyverse)
#library(magrittr)
knitr::opts_knit$set(root.dir = "Y:/users/rah/Teaching/2018fa_fw599")
```

```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/trainData_Oxyura_jamaicensis.RData") # duckData
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/testData_Oxyura_jamaicensis.RData") # duckData
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/testSubset_Oxyura_jamaicensis.RData") # duckData
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v3.RData") # envStack
```

```{r}
spp = "Oxyura_jamaicensis"
sppCol = which(names(trainData)==spp)

# Set up indices that may be useful later
trainPresIdxs = which(trainData[,sppCol]!=0)
trainAbsIdxs = which(trainData[,sppCol]==0)
testPresIdxs = which(testData[,sppCol]!=0)
testAbsIdxs = which(testData[,sppCol]==0)
testSubsetPresIdxs = which(testSubset[,sppCol]!=0)
testSubsetAbsIdxs = which(testSubset[,sppCol]==0)

```

# Decision: Which metrics should we compute? For presence/absence? For abundance?

Proposal: make PR and ROC curves for presence/absence data, and compute area under the curves. For abundance, we can use root mean squared error, but I suspect it will be dominated by the largest counts. I'm also interested in checking a less stringent criterion: whether the predicted abundance is the same order of magnitude as the true value (i.e., 0, 1-9, 10-99, 100-999, 1000-9999, >10000). I'm open to more ideas on this. We can also truncate or rescale the abundance predictions to presence/absence to compare directly to the presence/absence models. 

# Null models

We can use the geographic null model (inverse distance) of Hijmans (2012), implemented in the dismo package. This prediction takes a long time.
```{r,eval=FALSE}
nullModel = geoDist(cbind(trainData$LONGITUDE[trainPresIdxs],
                          trainData$LATITUDE[trainPresIdxs]),
                    lonlat=TRUE)
nullPred = predict(nullModel,envStack[[1]])
save(nullPred,file = "geoNullPreds.RData")
load(file = "geoNullPreds.RData")
plot(nullPred)
# next: retrieve test predictions from raster
geoNullPreds = extract(x=nullPred,
                y=cbind(testData$LONGITUDE,
                          testData$LATITUDE))
subsetGeoNullPreds = extract(x=nullPred,
                y=cbind(testSubset$LONGITUDE,
                          testSubset$LATITUDE))
```

Set up more basic null models as well:
```{r}
# random probabilities
paNullPreds = as.vector(array(runif(dim(testData)[1]),
                              c(dim(testData)[1],1)))
paSubsetNullPreds = as.vector(array(runif(dim(testData)[1]),
                                    c(dim(testSubset)[1],1)))

# permuted abundance values
abundNullPreds = sample(testData$abund)
abundSubsetNullPreds = sample(testSubset$abund)
```

## Presence/Absence
AUROCs on whole test set when predicting random probabilities for all test points:
```{r}
ROCcurveTestPAgeo = roc(response=testData$presAbs,
                     predictor=geoNullPreds)
ROCcurveSubsetPAgeo = roc(response=testSubset$presAbs,
                     predictor=subsetGeoNullPreds)
ROCcurveTestPA = roc(response=testData$presAbs,
                     predictor=paNullPreds)
ROCcurveSubsetPA = roc(response=testSubset$presAbs,
                       predictor=paSubsetNullPreds)
ROCcurveTestAbund = roc(response=testData$presAbs,
                     predictor=abundNullPreds)
ROCcurveSubsetAbund = roc(response=testSubset$presAbs,
                       predictor=abundSubsetNullPreds)

plot(ROCcurveTestPA) # plot it
plot(ROCcurveSubsetPA,col="red",add=TRUE)
plot(ROCcurveTestAbund,col="green",add=TRUE)
plot(ROCcurveSubsetAbund,col="blue",add=TRUE)
plot(ROCcurveTestPAgeo,col="orange",add=TRUE)
plot(ROCcurveSubsetPAgeo,col="purple",add=TRUE)
```

AUPRCs:
```{r}
PRCurveTestGeo = pr.curve(geoNullPreds,
                         weights.class0=testData$presAbs,
                         curve=TRUE)
PRCurveSubsetGeo = pr.curve(subsetGeoNullPreds,
                         weights.class0=testSubset$presAbs,
                         curve=TRUE)
PRCurveTestPA = pr.curve(paNullPreds,
                         weights.class0=testData$presAbs,
                         curve=TRUE)
PRCurveSubsetPA = pr.curve(paSubsetNullPreds,
                         weights.class0=testSubset$presAbs,
                         curve=TRUE)
PRCurveTestAbund = pr.curve(abundNullPreds,
                         weights.class0=testData$presAbs,
                         curve=TRUE)
PRCurveSubsetAbund = pr.curve(abundSubsetNullPreds,
                         weights.class0=testSubset$presAbs,
                         curve=TRUE)
plot(PRCurveTestPA,col="black")
plot(PRCurveSubsetPA,col="red",add=TRUE)
plot(PRCurveTestAbund,col="green",add=TRUE)
plot(PRCurveSubsetAbund,col="blue",add=TRUE)
plot(PRCurveTestGeo,col="orange",add=TRUE)
plot(PRCurveSubsetGeo,col="purple",add=TRUE)

```

## Abundance

RMSE on whole test set and subset when predicting permuted abundance:
```{r}
testSqErr = (abundNullPreds - testData$abund)^2
table(testSqErr)
sqrt(mean(testSqErr))
```

```{r}
testSubsetSqErr = (abundSubsetNullPreds - testSubset$abund)^2
table(testSubsetSqErr)
sqrt(mean(testSubsetSqErr))
```

How about order of magnitude? Here for the whole test set.
```{r}
abundTestPredsOM = abundNullPreds
abundTestPredsOM[which(abundTestPredsOM>=1 & abundTestPredsOM<10)] = 1
abundTestPredsOM[which(abundTestPredsOM>=10 & abundTestPredsOM<100)] = 2
abundTestPredsOM[which(abundTestPredsOM>=100 & abundTestPredsOM<1000)] = 3
abundTestPredsOM[which(abundTestPredsOM>=1000 & abundTestPredsOM<10000)] = 4
abundTestPredsOM[which(abundTestPredsOM>=10000)] = 5

testAbundOM = testData$abund
testAbundOM[which(testAbundOM>=1 & testAbundOM<10)] = 1
testAbundOM[which(testAbundOM>=10 & testAbundOM<100)] = 2
testAbundOM[which(testAbundOM>=100 & testAbundOM<1000)] = 3
testAbundOM[which(testAbundOM>=1000 & testAbundOM<10000)] = 4
testAbundOM[which(testAbundOM>=10000)] = 5

CMtest = table(testAbundOM,abundTestPredsOM)
CMtest
print(paste("Proportion of test points on the diagonal:",sum(diag(CMtest))/sum(CMtest)))
```

Order of magnitude for test subset.
```{r}
abundSubsetPredsOM = abundSubsetNullPreds
abundSubsetPredsOM[which(abundSubsetPredsOM>=1 & abundSubsetPredsOM<10)] = 1
abundSubsetPredsOM[which(abundSubsetPredsOM>=10 & abundSubsetPredsOM<100)] = 2
abundSubsetPredsOM[which(abundSubsetPredsOM>=100 & abundSubsetPredsOM<1000)] = 3
abundSubsetPredsOM[which(abundSubsetPredsOM>=1000 & abundSubsetPredsOM<10000)] = 4
abundSubsetPredsOM[which(abundSubsetPredsOM>=10000)] = 5

testSubsetAbundOM = testSubset$abund
testSubsetAbundOM[which(testSubsetAbundOM>=1 & testSubsetAbundOM<10)] = 1
testSubsetAbundOM[which(testSubsetAbundOM>=10 & testSubsetAbundOM<100)] = 2
testSubsetAbundOM[which(testSubsetAbundOM>=100 & testSubsetAbundOM<1000)] = 3
testSubsetAbundOM[which(testSubsetAbundOM>=1000 & testSubsetAbundOM<10000)] = 4
testSubsetAbundOM[which(testSubsetAbundOM>=10000)] = 5

CMsubset = table(testSubsetAbundOM,abundSubsetPredsOM)
CMsubset
print(paste("Proportion of test points on the diagonal:",sum(diag(CMsubset))/sum(CMsubset)))
```

## A spatial look at residuals
Let's also plan to plot the model residuals spatially and look at their autocorrelation.

