---
title: "Merge rasters"
output: html_notebook
---

This document contains code to add layers to the stack of predictor variables.

```{r,message=FALSE}
source("Y:/users/rah/R/Rconfigure_default.r")
knitr::opts_knit$set(root.dir = "Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/")

# geographic tools packages
library(raster)
library(rgdal)
```

Load the current version of envStack and a new elevation layer. Crop the elevation layer to the extent of envStack, and then add it to the stack. Save the new version.
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack.RData")

elevation <- raster('C:/Users/Rebecca/Desktop/SDM_Workshop/WORLDCLIM/Elevation/World/elevation.tif')
elevation <- resample(elevation,envStack$Annual.Mean.Temperature)
#elevation <- crop(elevation, extent(envStack))
envStack_merged = addLayer(envStack,elevation)
#save(envStack_v2,"Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2.RData")
```

```{r}
landsat <- stack('L8_surfaceReflectance_tnc_terr_ecoregions_7bands_1500m.tif')
res(landsat)
res(envStack)
```

```{r}
flyways <- raster("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/DU_raw/flyraster.ovr")
focusAreas <- raster("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/DU_raw/focusraster.ovr")
plan <- raster("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/DU_raw/planraster.ovr")

```


Ray Brunner: wetlands layer
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2RLB.Rdata") # envStack_v2

envStack_merged = addLayer(envStack_merged,envStack_v2$glwd_3)

```

Hannah Friedrich: GlobCover
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_HF.RData") # envStack_v4

envStack_merged = addLayer(envStack_merged,envStack_v4$GlobCover2009LC)

```

Francisco Pickens: MODIS
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_FATP.Rdata") # land_usetype

envStack_merged = addLayer(envStack_merged,land_usetype$LC_5min_global_2012)

```

Marisa Zimmerman: Ducks Unlimited (flyways?)
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2MZ.RData") # envStack3


```

```{r}
envStack = envStack_merged
save(envStack,file="Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2.RData")
```


Add "standard observer" values for spatial prediction with effort variables. 
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v4.RData")
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/janDuckDataNAmer_v2.RData")

dayRast = envStack$Annual.Mean.Temperature
dayRast = dayRast*0+16 # 1/16
names(dayRast) = "DAY"
envStack = addLayer(dayRast,envStack)

timeRast = envStack$Annual.Mean.Temperature
timeRast = timeRast*0+8 # 8am
names(timeRast) = "TIME"
envStack = addLayer(timeRast,envStack)

typeRast = envStack$Annual.Mean.Temperature
typeRast = typeRast*0 + 1 # first count type P21 (stationary)
typeRast = as.factor(typeRast)
#typeLevs = data.frame(ID=1:nlevels(duckData$COUNT_TYPE))
#typeLevs$Name = levels(duckData$COUNT_TYPE)
typeLevs = levels(typeRast)[[1]]
typeLevs$code = "P21"
levels(typeRast) = typeLevs
names(typeRast) = "COUNT_TYPE"
envStack = addLayer(typeRast,envStack)

hoursRast = envStack$Annual.Mean.Temperature
hoursRast = hoursRast*0 + 1 # one hour
names(hoursRast) = "EFFORT_HRS"
envStack = addLayer(hoursRast,envStack)

distRast = envStack$Annual.Mean.Temperature
distRast = distRast*0
names(distRast) = "EFFORT_DISTANCE_KM"
envStack = addLayer(distRast,envStack)

areaRast = envStack$Annual.Mean.Temperature
areaRast = areaRast*0
names(areaRast) = "EFFORT_AREA_HA"
envStack = addLayer(areaRast,envStack)

numRast = envStack$Annual.Mean.Temperature
numRast = numRast*0 + 1
names(numRast) = "NUMBER_OBSERVERS"
envStack = addLayer(numRast,envStack)

save(envStack,file="Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v5.RData")


```