---
title: "Merge rasters"
output: html_notebook
---

This document contains code to add layers to the stack of predictor variables.

```{r,message=FALSE}
source("Y:/users/rah/R/Rconfigure_default.r")
knitr::opts_knit$set(root.dir = "Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/")

# geographic tools packages
library(raster)
library(rgdal)
```

Load the current version of envStack and a new elevation layer. Crop the elevation layer to the extent of envStack, and then add it to the stack. Save the new version.
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack.RData")

elevation <- raster('C:/Users/Rebecca/Desktop/SDM_Workshop/WORLDCLIM/Elevation/World/elevation.tif')
elevation <- resample(elevation,envStack$Annual.Mean.Temperature)
#elevation <- crop(elevation, extent(envStack))
envStack_merged = addLayer(envStack,elevation)
#save(envStack_v2,"Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2.RData")
```

```{r}
landsat <- stack('L8_surfaceReflectance_tnc_terr_ecoregions_7bands_1500m.tif')
res(landsat)
res(envStack)
```

```{r}
flyways <- raster("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/DU_raw/flyraster.ovr")
focusAreas <- raster("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/DU_raw/focusraster.ovr")
plan <- raster("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/DU_raw/planraster.ovr")

```


Ray Brunner: wetlands layer
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2RLB.Rdata") # envStack_v2

envStack_merged = addLayer(envStack_merged,envStack_v2$glwd_3)

```

Hannah Friedrich: GlobCover
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_HF.RData") # envStack_v4

envStack_merged = addLayer(envStack_merged,envStack_v4$GlobCover2009LC)

```

Francisco Pickens: MODIS
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_FATP.Rdata") # land_usetype

envStack_merged = addLayer(envStack_merged,land_usetype$LC_5min_global_2012)

```

Marisa Zimmerman: Ducks Unlimited (flyways?)
```{r}
load("Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2MZ.RData") # envStack3


```

```{r}
envStack = envStack_merged
save(envStack,file="Y:/users/rah/Teaching/2018fa_fw599/GroupProject/FW599/envStack_v2.RData")
```
